---
title: "Demystifying Hierarhcical Models"
author: Jeff Rouder
date:  May, 2024

output:
  beamer_presentation:
    theme: "metropolis"
    fonttheme: "structurebold"
    fig_caption: false
    incremental: false
    
header-includes   :
   - \usepackage{bm}
   - \usepackage{pcl}
   - \usepackage{amsmath}    
---


## Hierarchical Models Seem Magical

\[\begin{aligned}
Y_{ij} &\sim \mbox{N}(\mu_i,\sigma^2)\\
\mu_i &\sim \mbox{N}(\nu,\delta)
\end{aligned}
\]

is the same as:

\[\begin{aligned}
Y_{ij} &\sim \mbox{N}(\nu+\mu_i,\sigma^2)\\
\mu_i &\sim \mbox{N}(0,\delta)
\end{aligned}
\]

More standard:
\[\begin{aligned}
Y_{ij} &\sim \mbox{N}(\mu+\mu_i,\sigma^2)\\
\alpha_i &\sim \mbox{N}(0,\delta)
\end{aligned}
\]

If there are two people, there are three parameters, and that seems like magic!

## What Happens With 2 People and 3 Parameters?

```{R,echo=T}
source('jags/normJ.R')
```

## What Happens With 2 People and 3 Parameters?
```{R,echo=T}
sub=rep(1:2,each=100)
tMu=c(8,12)
y=rnorm(200,tMu[sub],5)
dat=data.frame(sub,y)
prior=list("a"=10,"b"=10^2,'r1'=10^2,'r2'=10^2)
out=runM2a(dat,prior)
alpha=out$BUGSoutput$sims.list$alpha
mu=out$BUGSoutput$sims.list$mu
```

## How Well Should We Be Able To Localize Parameter?

SEM = $\sigma/\sqrt{n}$ = 5/10 =.5

We should know all parameter $\pm 1$.

## Yet, Can't Localize Parameters Well

```{r,echo=T}
hist(mu)
```

## Can't Localize Parameters Well

```{r,echo=T}
hist(alpha[,1])
```

## Correlated Parameters ...

```{r}
par(mfrow=c(1,2))
plot(mu,alpha[,1])
plot(alpha)
```
## What is estimable?

```{r}
par(mfrow=c(1,2))
hist(mu+alpha[,2])
hist(alpha[,2]-alpha[,1])
```

## Morale of the Story

There is no magic here.  Usually, it matters little how we parameterized the hierarchical models because the cost of one parameter is not great if there are a lot of people.  






